---
title: "Practice 11: SARIMA and ARFIMA models"
author: "Carlo Alessi"
output:
  pdf_document:
    fig_caption: yes
    number_sections: true
---

\section{Fitting SARIMA models}

\paragraph{Example 1}
In this example we work with the time series of monthly temperatures at Nottingham Castle, in degrees Farenheit, from 1921 to 1940. Figure \ref{fig:not_series_acf} shows the time series and the autocorrelation function.

```{r fig.cap="Nottingham Castle time series and autocorrelation function.\\label{fig:not_series_acf}"}
library(forecast)
layout(1:2)
x <- nottem
plot(x)
acf(x)
```

Figure \ref{fig:nottingham_12} shows the Nottingham time series after differencing once with lag 12 $(Id-B_{12})X$. Autocorrelation and partial autocorrelation functions are also shown.
```{r fig.cap="Nottingham time series of first differences with lag 12, autocorrelation and partial autocorrelation function.\\label{fig:nottingham_12}"}
layout(1:3)
y <- diff(x,12)
plot(y, ylab="dy_12")

acf(y)
pacf(y)
```

Fit the differenced series with an ARMA(1,1) model. In \autoref{fig:residuals} are plotted the residuals of the model used to fit the series in figure \ref{fig:nottingham_12}.

```{r fig.cap="Residuals of fitted ARMA(1,1) model.\\label{fig:residuals}"}
fit1 <- arima(y,order=c(1,0,1), seasonal=list(order=c(1,0,1), period=12))
fit1

fit2 = resid(fit1)
plot(fit2,  ylab="residuals")
```

Figure \ref{fig:acf_pacf_res} show the autocorrelation and partial autocorrelation functions of the residuals.
```{r fig.cap="Autocorrelation and partial autocorrelation function of the residuals.\\label{fig:acf_pacf_res}"}
par(mfrow=c(2,1))
acf(fit2, ylab="ACF residuals")
pacf(fit2, ylab="PACF residuals")
```

According to the Ljung-Box test, the residuals are an IID noise. Figure \ref{fig:nottingham_forecasting} shows the time series forecasting 24 steps ahead.
```{r fig.cap="Nottingham time series forcasting.\\label{fig:nottingham_forecasting}"}
Box.test(fit2, log(240), type="Ljung-Box")
predict(fit1,n.ahead=24)

xfc <- forecast(x)
plot(xfc, xlab="time", ylab="x")
```

\paragraph{Example 2}
In this example we work with the AirPassengers time series (see figure \ref{fig:passengers}). 
Three $SARIMA(p,d,q)(a,D,c)_s$ models are used to fit the time series:

\begin{itemize}
\item[1.] $SARIMA(0,1,0)(0,1,0)_{12}$
\item[2.] $SARIMA(1,1,0)(0,1,0)_{12}$
\item[3.] $SARIMA(0,1,0)(0,1,1)_{12}$
\end{itemize}

The 3 SARIMA models are shown in respectively in figure \ref{fig:sarima_1}, \ref{fig:sarima_2}, and \ref{fig:sarima_3}. The best performing model is the second one, because the p-values relative to the IID noise hypothesis test are high for many lags.

```{r fig.cap="AirlinePassengers time series.\\label{fig:passengers}"}
x<-AirPassengers
plot(x)
```

```{r fig.cap="SARIMA(0,1,0,0,1,0,12).\\label{fig:sarima_1}"}
library(astsa)
fit1<-sarima(x,0,1,0,0,1,0,12)
```

```{r fig.cap="SARIMA(1,1,0,0,1,0,12).\\label{fig:sarima_2}"}
fit2<-sarima(x,1,1,0,0,1,0,12)
```

```{r fig.cap="SARIMA(0,1,0,0,1,1,12).\\label{fig:sarima_3}"}
fit3<-sarima(x,0,1,0,0,1,1,12)
```

The following code forecasts the AirlinePassengers time series 12 steps ahead using a $SARIMA(0,1,0)(0,1,1)_{12}$ model. See figure \ref{fig:airline_forecast}.
```{r fig.cap="Airline passenger time series forecasting.\\label{fig:airline_forecast}"}
ahead = 12
seasonal_period = 12
sarima.for(x, n.ahead = ahead, 0, 1, 0, 0, 1, 1, S=seasonal_period)
```

\section{Dickey-Fuller test}
Dickey-Fuller test allows to distinguish between a random walk (not stationary) and an AR(1) model (stationary) when parameter $\phi$ is near to 1. The null hypothesis is $\phi=1$.
The p-value is very small so we reject the hypothesis of random walk.
```{r}
library(tseries)
x<-rnorm(200) # simulate a white noise
adf.test(x)
```

The following is a random walk (see figure \ref{fig:random_walk}). The p-value is big and so, we accept the hypothesis that y is a random walk, or equivalently, $\phi=1$. Therefore, it is convenient to take first differences to adjust an ARMA model. 
```{r fig.cap="Random walk.\\label{fig:random_walk}"}
y <- cumsum(x)
ts.plot(y)
# test
adf.test(y)
```

\paragraph{Example 3}
Simulate and AR(1) model with $\phi=0.99$, which is plotted in figure \ref{fig:ar_1}. Check what happens.
```{r fig.cap="AR(1) model.\\label{fig:ar_1}"}
n<-2000
p<-0.99
x<-arima.sim(list(order=c(1,0,0), ar=p),n)
plot(x,type="l")
adf.test(x)
```

The following code fits the time series with and ARFIMA model with $p=2$ and $q=1$, maximizing the likelihood estimators of the parameters $(p,d,q)$ in the range $[0, 0.5)$. Then the time series is first  differenced (fit2), then fitted with and ARMA(2,1) model (fit3). The rediduals of the resulted time series (fit4) is an IID noise, as reported in the output of the code. The autocorrelation functions of fit2 and fit4 are plotted in figure \ref{fig:fit_24}.
```{r fig.cap="Autocorrelation function of differenced series (fit2, above) and residuals (fit4, below).\\label{fig:fit_24}"}
library(fracdiff)
layout(1:2)

fit1 <- fracdiff(x, nar=2, nma=1, drange=c(0,0.5))
d <- fit1$d  # retrieve 'd' value

fit2 = diffseries(x,d)
acf(fit2)

fit3 <- arima(fit2, order=c(2,0,1))

fit4 <- resid(fit3)
acf(fit4)
Box.test(fit4)
```

\paragraph{Example}
Simulate a long-memory time series from and fractional ARIMA(1,1,0) with $\phi=0.9$ and $d=0.4$ (see figure \ref{fig:arfima}). The time series is then fitted with an ARFIMA model with $p=1$.
```{r fig.cap="ARIMA(1,1,0) with $\\phi=0.9$ and $d=0.4$. \\label{fig:arfima}"}
library(fracdiff)
n<-1000
sample <- fracdiff.sim(n, ar=0.9, d=0.4)
x <- sample$series
plot(x,type="l")

fit1 <- fracdiff(x,nar=1)
summary(fit1)
```
The time series is differenced with fractional degree $d$. The resulted series is fit with the best AR(1) model, which residuals are an IID noise. The autocorrelation functions for the differenced time series and the residuals are shown in figure \ref{fig:diff_res}.
```{r fig.cap="Autocorrelation function of the differenced time series (top) and of the residuals (bottom).\\label{diff_res}"}
d<-fit1$d
y1<-diffseries(x,d)
acf(y1)

y2<-arima(y1,order=c(1,0,0))
z<-resid(y2)
acf(z)
Box.test(z,log(n),type=c("Ljung-Box"))
```

\paragraph{Example on the Australian residents time series}
The same method is applied to the time series of quarterly data of Australian residents from 1972 to 1993.
Autocorrelation functions of $x$ and $dx$ are shown in figure \ref{fig:acf_x_dx}.

```{r fig.cap="Autocorrelation funtion of time series x and dx. \\label{fig:acf_x_dx}"}
library("datasets")
layout(1:2)
x<-austres
acf(x, ylab="ACF x")
dx<-diff(x)
acf(dx, ylab="ACF dx")
```

Estimate parameter $d$, and differentiate the time series $dx$ obtaining the residuals. Center the residuals by subtracting the mean. The IID noise hypothesis on the centred residuals is accepted. The autocorrelation function of the centred residuals is plotted in figure \ref{fig:acf_0_res}.

```{r fig.cap="Autocorrelation function of centred residuals.\\label{fig:acf_0_res}"}
fdx <- fracdiff(dx, nar=0, nma=0, drange=c(0,0.5))
summary(fdx)
d = fdx$d
d
resid <- diffseries(dx,d)
y <- resid-mean(resid)
acf(y)
Box.test(y, lag=log(88), type=c("Ljung-Box"))
```

