---
title: 'Homework 5: AR(p) models'
author: "Carlo Alessi"
output:
  pdf_document:
    fig_caption: yes
    number_sections: true
---
\section{Simulating AR(p) models}
The code chunks below simulate several time series from an AR(2) model, and compute their autocorrelation function. Figure \ref{fig:ar2_corr_1} shows a time series of size $n=1000$ generated from an AR(2) model with parameters $\phi_1 = 0.8897$ and $\phi_2 = - 0.4858$.

```{r fig.cap="AR(2) model with parameters $(\\phi_1, \\phi_2) = (0.8897, -0.4858)$ and its correlogram.\\label{fig:ar2_corr_1}"}
layout(2:1)
n <- 1000
x <- arima.sim(list(ar = c(0.8897, -0.4858), sd = sqrt(0.1796)), n)
plot(x,type="l")
acf(x)
```


Figure \ref{fig:ar2_corr_2} shows a time series of size $n=100$ generated from an AR(2) model with parameters $\phi_1 = 0.5$ and $\phi_2 = - 0.5$.
```{r fig.cap="AR(2) model with paramenters $(\\phi_1, \\phi_2) = (0.5, -0.5)$ and its correlogram.\\label{fig:ar2_corr_2}"}
layout(2:1)
n <- 100
y<-arima.sim(list(ar = c(0.5, -0.5), sd = sqrt(3)), n)
plot(y,type="l")
acf(y)
```

Figure \ref{fig:ar2_corr_3} shows a time series of size $n=500$ generated from an AR(2) model with parameters $\phi_1 = 0.7$ and $\phi_2 = - 0.2$.
```{r fig.cap="AR(2) model with paramenters $(\\phi_1, \\phi_2) = (0.7, -0.2)$ and its correlogram.\\label{fig:ar2_corr_3}"}
layout(2:1)
n <- 500
z <- arima.sim(list(ar = c(0.7, -0.2), sd = sqrt(2)), n)
plot(z,type="l")
acf(z)
```

\section{Fitting AR(p) models}
The following code chunks fit the time series generated in the previous section (x,y,z) with an AR(2) model using Maximum Likelihood.
```{r}
s1<-arima(x, order=c(2,0,0), method="ML")
s1
```

```{r}
s2<-arima(y, order=c(2,0,0), method="ML")
s2
```

```{r}
s3<-arima(z, order=c(2,0,0), method="ML")
s3
```

The following code chunks give an alternative method to fit an autoregressive time series using Maximum Likelihood Estimation. In the first chunk the data is fit using a second order model, whereas in the second chunk it is used a third order model. Since a second order model is sufficient, the results do not change.
```{r}
ar(x, order.max=2, method="mle")
```
```{r}
ar(x, order.max=3, method="mle")
```

The following code chunk simulate an ARIMA model with AR and MA coefficients respectively -0.6 and 0.5 (see figure \ref{fig:arima}).
```{r fig.cap="ARIMA model with coefficients $(ar, ma) = (-0.6, -0.5)$.\\label{fig:arima}"}
n<-10000
x<-arima.sim(list(ar=-0.6, ma=0.5), n)
plot(x,type="l")
coef(arima(x,order=c(1,0,1)))
```
\section{Validation}
In the following code chunk it is applied the Ljung-Box test to the time series shown in figure \ref{fig:arima}.
It can be seen that the IID noise hypothesis has been rejected since the p-value is less than 0.05.
```{r}
h<-floor(log(n))
Box.test(x, lag=h, type=c("Ljung-Box"))
```
In the following code chunks we apply the Ljung-Box test to a time series $X$ simulated from an AR(2) model with parameters $(\phi_1, \phi_2) = (0.8897, - 0.4858)$ (see figure \ref{fig:ar2_corr_1}). Then the test is applied to the residuals of the AR(2) model used to fit $X$. The IID noise hypothesis is accepted for the residuals and rejected on the original time series.
```{r}
n <- 1000
x <- arima.sim(list(ar = c(0.8897, -0.4858), sd = sqrt(0.1796)), n)
#plot(x,type="l")

s <- arima(x, order=c(2,0,0), method="ML")
s

y<-resid(s) # extract residuals from fitted model

h<-floor(log(n))
Box.test(x,lag=h, type=c("Ljung-Box"))
Box.test(y,lag=h,type=c("Ljung-Box"))
```

\section{Selection of the best AR(p) model}
The instruction selects the best AR between AR(1), AR(2) and AR(3) to fit data x (see figure \ref{fig:ar2_corr_1}). The solution is an AR(2).
```{r fig.cap="AR"}
n<-1000
x<-arima.sim(list(ar = c(0.8897, -0.4858), sd = sqrt(0.1796)), n)
#plot(x,type="l")
ss1 <- ar(x, order.max=3, method="mle")
ss1
```

\section{A complete analysis with real data}
The following time series reppresents annual measurements of the Lake Huron level (in feet). The graph of the time series and its correolagram is shown in figure \ref{fig:lake}.
```{r}
x <- LakeHuron
x
```

```{r fig.cap="Annual measurements of the Lake Luron level (in feet).\\label{fig:lake}"}
layout(2:1)
plot (x, ylab="depth", xlab="time")
acf(x)
```

In figure \ref{fig:lake_lag} are shown plots of the Lake Huron time series against lagged version of itself (e.g $(x_{i-l}, x_i)$ for lags $l = 1,...,4$). As the lag increases the linear trend disappears.
```{r fig.cap="Time series lag plot of Lake Huron data. \\label{fig:lake_lag}"}
lag.plot (x, lag=4, do.lines=F)
```



Then we fit an AR(1) model to the time series using the Yule-Walker method.
```{r}
s1<-ar(x, order.max=1, method="yw")
s1
```

The following code chunks fit an AR(p) model to the time series using the Maximum Likelihood Estimation and the Yule-Walker. Both methods propose an AR(2) model with similar parameter.
```{r}
s3<-ar(x, method="mle")
s3
```

```{r}
s4<-ar(x, method="yw")
s4
```

