---
title: 'Practice 12: GARCH models'
author: "Carlo Alessi"
output:
  html_document: default
  pdf_document:
    fig_caption: yes
    number_sections: yes
---

\section{Nonlinear correlation and GARCH models}
In this section it is analyzed the daily returns of the S\&P500 time series.
From the Ljung-Box test we see that it is a white noise, since the p-value is greater than 0.05. The series and its ACF are shown in figure \ref{sp500}.
```{r fig.cap="Plot of the SP500 time series and its ACF.\\label{sp500}"}
library(MASS)
data(SP500)

layout(1:2)
plot(SP500, type="l")
acf(SP500)
Box.test(SP500, type="Ljung-Box")
```

The following code subtracts the mean from the SP500 time series, obtaining the series $X-\mu$. The series is still a white noise. This can be seen from the ACF in figure \ref{centredsp500}.

```{r, fig.cap="The zero-centred SP500 time series and its ACF.\\label{centredsp500}"}
layout(1:2)
m<-mean(SP500)
x<-SP500-m
plot(x, type="l")
acf(x)
```

But squaring the centred series ($(X-\mu)^2$) makes the IID noise hypothesis to be rejected, since the $p\-value < 0.05$. This is also confirmed by the ACF in figure \ref{squared}, which shows that there are several significant correlations at different lags.

```{r fig.cap="ACF of the $(X-\\mu)^2$.\\label{squared}"}
y<-x^2
acf(y)
Box.test(y, type="Ljung-Box")
```

\section{Example of ARCH(1)}
The following code block manually generates a time series of length $n=9999$ from an ARCH(1) model with parameters $\alpha_0 = 0.25$ and $\alpha_1 = 0.5$. The time series is shown in figure \ref{arch1model}.

```{r fig.cap="ARCH(1) model with $\\alpha_0 = 0.25$ and $\\alpha_1 = 0.5$.\\label{arch1model}"}
a0 <- 0.25
a1 <- 0.5

n = 9999
x <- numeric(n)
x0 <- rnorm(1)

for (i in 1:n){
  x0 <- rnorm(1, sd=sqrt(a0 + a1 * x0^2))
  x[i] <- x0
}

plot(x, type="l")

```

\section{Example of GARCH(1,1)}
The code below shows an alternative example for a GARCH(1, 1), with parameters $\alpha_0 = 0.1$, $\alpha_1 = 0.4$ and $\beta_1 = 0.2$. The generated time series $Y$ is shown in figure \ref{garch11} along with its ACF, and the ACF of $Y^2$.
We can see that $Y^2$ is not an iid noise.
 
```{r fig.cap="GARCH(1,1) model with $\\alpha_0 = 0.1$, $\\alpha_1 = 0.4$ and $\\beta_1 = 0.2$.\\label{garch11}"}
n <- 10000
a0 <- 0.1
a1 <- 0.4
b1 <- 0.2
w <- rnorm(n)
y <- rep(0,n)
h <- rep(0,n)

for(i in 2:n){
  h[i] <- a0 + a1 * (y[i-1]^2) + b1 * h[i-1]
  y[i] <- w[i] * sqrt(h[i])
}

layout(1:3)
plot(y, type="l")
acf(y, ylab="ACF y") 
acf(y^2, ylab="ACF y^2")
```

\paragraph{Squaring the time series}
The time series $X$ in the example related to figure \ref{arch1model} is a white noise, but $X^2$ is an AR(1) process. Figure \ref{xsquared} shows the squared time series, and compares its correlogram with the one of the original time series.

```{r fig.cap="Plot of squared time series $X^2$, ACF of $X$ and ACF of $X^2$.\\label{xsquared}"}
layout(1:3)
y <- x^2
plot(y, type="l")
acf(x, ylab="ACF x")
acf(y, ylab="ACF x^2")
```

\paragraph{Testing the IID hypothesis}
Performing the Ljung-Box test on $X$ and $Y=X^2$ we see that the former is a white noise, while the latter is not.
```{r}
Box.test(x, lag = 10, type = c("Ljung-Box"))
Box.test(y, lag = 10, type=c("Ljung-Box"))
```

\section{Fitting GARCH models}

The following code chunk fits to the time series $X$ with the following models:
\begin{itemize}
\item ARCH(1)
\item ARCH(2)
\item GARCH(1,1)
\end{itemize}

The output of the code shows the parameters estimates for each model.

```{r}
library(tseries)
garch(x,order=c(0,1),trace=F)
garch(x,order=c(0,2),trace=F)
garch(x,order=c(1,1),trace=F)
```

