---
title: "Practice 6: Simulating and fitting ARIMA models"
author: "Carlo Alessi"
output:
  pdf_document:
    fig_caption: yes
    number_sections: true
---
The goal of this practice is to identify a possible ARIMA model to a given data, to fit it and to validate it as a good model.

\section{Simulation of a model}
In the following code I simulate 3 time series respectively from:
\begin{itemize}
\item AR(1) model with parameter $\phi=0.7$, 
\item MA(2) model with parameters $\theta_1 = – 1.5$ and $\theta_2 = – 0.75$, 
\item ARMA(2,1) model with parameters $(\phi_1,\phi_2) = (0.3,0.2)$ and $\theta = – 0.3$.
\end{itemize}

```{r}
n = 10000
SIM1<-arima.sim(list(order=c(1,0,0), ar=0.7),n)
SIM2<-arima.sim(list(order=c(0,0,2), ma=c(1.5,0.75)),n)
SIM3<-arima.sim(list(order=c(2,0,1), ar=c(0.3,0.2), ma=0.3), n)
```

The plots of the time series generated above are shown in figure \ref{fig:time_series}.

```{r, fig.cap="Time series SIM1, SIM2, SIM3.\\label{fig:time_series}"}
layout(1:3)
plot(SIM1, type="l")
plot(SIM2, type="l")
plot(SIM3, type="l")
```

\section{Heuristic identification of the model. Auto-correlation and partial auto-correlation functions}
The autocorrelation and partial correlation functions of the time series SIM1, SIM2 and SIM3 are shown respectively in figure \ref{fig:auto_sim1}, \ref{fig:auto_sim2} and \ref{fig:auto_sim3}.

```{r fig.cap="Autocorrelation and partial autocorrelation function for SIM1.\\label{fig:auto_sim1}"}
layout(2:1)
k = log(n)
acf(SIM1, k)
pacf(SIM1, k)
```
```{r fig.cap="Autocorrelation and partial autocorrelation function for SIM2.\\label{fig:auto_sim2}"}
layout(2:1)
k = log(n)
acf(SIM2, k)
pacf(SIM2, k)
```
```{r fig.cap="Autocorrelation and partial autocorrelation function for SIM3.\\label{fig:auto_sim3}"}
layout(2:1)
k = log(n)
acf(SIM3, k)
pacf(SIM3, k)
```

\section{Fitting a model}
To determine the parameters $p$ and $q$ of an $ARMA(p,q)$ model we can look at the plots of the autocorrelation function (ACF) and partial autocorrelation function (PACF). A good value of $p$ is the maximum lag at which the PACF is significant, whereas a good value of $q$ is the maximum lag at which the ACF is significant.

In the code chunks below the time serieses SIM1, SIM2 and SIM3 are fitted respectively with ARMA(1,8), ARMA(0,2) and ARMA(1,4) models.

```{r}
S1 <- arima(SIM1, order=c(1,0,8), method="ML")
S1
```

```{r}
S2 <- arima(SIM2, order=c(0,0,2), method="ML")
S2
```

```{r}
S3 <- arima(SIM3, order=c(1,0,4), method="ML")
S3
```

\section{Validation}
First we apply the Ljung-Box test to the serieses SIM1, SIM2 and SIM3. The IID noise hypothesis is rejected in all of them. This suggests that the serieses are stationary. 
Then to validate the fitted models as good models the residuals must be IID noise. The Ljung-Box applied to the residuals gives a p-value greater than 0.05 for all models.

```{r}
Box.test (SIM1, lag = log(n), type = c("Ljung-Box"))
Box.test (SIM2, lag = log(n), type = c("Ljung-Box"))
Box.test (SIM3, lag = log(n), type = c("Ljung-Box"))
```

```{r}
y <- resid(S1)
Box.test(y, lag = log(n), type = c("Ljung-Box"))

y <- resid(S2)
Box.test(y, lag = log(n), type = c("Ljung-Box"))

y <- resid(S3)
Box.test(y, lag = log(n), type = c("Ljung-Box"))
```

\section{Example}
Figure \ref{fig:example} shows a complete example. The time series is generated from an ARMA(2,1) model with paramenters $\phi_1 =0.3$, $\phi_2 = 0.2$ and $\theta = - 0.3$. The time series is then fitted with a ARMA(2,1) model, whose residuals are IID noise.
```{r fig.cap="ARMA(2,1) model with parameters $(\\phi_1, \\phi_2) = (0.3,0.2)$ and $\\theta=-0.3$, autocorrelation and partial autocorrelation function.\\label{fig:example}"}
layout(3:1)
n<-10000
s3<-arima.sim(list(order=c(2,0,1),ar=c(0.3,0.2),ma=0.3), n)
plot(s3, type="l")

acf(s3, log(n))
pacf(s3, log(n))

p3<-arima(s3,order=c(2,0,1),method="ML")
p3
h<-log(n)

Box.test(s3,lag=h,type=c("Ljung-Box"))
y<-resid(p3)
Box.test(y,lag=h,type=c("Ljung-Box"))
```


